---
title: "EDA - Don River"
author: "Dylan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```
## Overview

This notebook explores the clean daily streamflow record for the Don 
River at Todmorden (Station 02HC024) covering 1961 to 2023. The 
goal is not to build anything yet, it is to understand what 
the data looks like, identify any quality issues, and connect 
observations directly to the modeling decisions made in later 
notebooks.

The analysis moves through four questions:

1. **Is the data clean and complete?**
missing values, quality flags, incomplete years, duplicates, and extreme values are all checked before any analysis begins

2. **What does the distribution look like?** 
the shape of the flow distribution directly determines which statistical model is appropriate for severity

3. **Are there seasonal patterns?**
flood frequency and severity vary by season, which has implications for the stationarity assumption in Version 1

4. **Is the series stationary?** 
the annual maximum trend and autocorrelation structure reveal whether the data supports a stationary model or motivates a time-varying extension

**Key outputs from this notebook:**

- Confirmation that the data is suitable for modeling
- Visual and statistical evidence supporting GPD for severity
- Visual and statistical evidence supporting Poisson for frequency
- Empirical justification for the 3 day declustering criterion
- Documented limitations of the stationarity assumption

All modeling decisions referenced here are fully justified in 
[METHODOLOGY.md](https://github.com/dylanthackeray/toronto_flood_cat_model/blob/main/METHODOLOGY.md). 

---

## 1. Load Data
```{r}
library(tidyverse)
library(lubridate)
library(here)

flow_data <- read_csv(here::here("data", "clean", "02HC024_clean.csv"))
```

---

## 2. Data Quality Checks

### Structure
```{r}
glimpse(flow_data)
head(flow_data)
```

### Missing Values & Data Quality
```{r}
# NA check across all columns
flow_data %>%
  summarise(across(everything(), ~ sum(is.na(.))))

# proportion of each quality flag
quality_summary <- flow_data %>%
  count(quality) %>%
  mutate(proportion = round(n / sum(n) * 100, 1))

quality_summary

measured  <- mean(flow_data$quality == "measured")
estimated <- mean(flow_data$quality == "estimated")
ice  <- mean(flow_data$quality == "ice_conditions")
partial <- mean(flow_data$quality == "partial_day")
```

The data is mostly clean: `r round(measured * 100, 1)`% of observations are measured values with no quality concern. The remaining `r round((1 - measured) * 100, 1)`% breaks down as follows: `r round(estimated * 100, 1)`% estimated (gap-filled), `r round(ice * 100, 1)`% ice-affected, and `r round(partial * 100, 1)`% partial day recordings.

Ice-affected measurements are worth noting for a flood frequency model as winter flow readings under ice conditions can be unreliable. However at `r round(ice * 100, 1)`% of total observations this is not affecting the analysis. All flagged observations are kept but their quality status is carried through as a variable for sensitivity checks if needed.

### Complete Years
```{r}
incomplete_years <- flow_data %>%
  group_by(year) %>%
  summarise(n_days = n()) %>%
  filter(n_days < 330)

incomplete_years
```

`r if(nrow(incomplete_years) == 0) "All years have at least 330 days of observations so no years flagged for removal." else paste(nrow(incomplete_years), "year(s) have fewer than 330 observations:", paste(incomplete_years$year, collapse = ", "), ". These are flagged for potential removal before modeling. The 330 day cutoff represents 90% of the year. The years below this threshold risk underrepresenting flood frequency due to missing observations rather than the absence of events.")`

### Date Range
```{r}
range(flow_data$date)
```

### Negative Or Zero Flows (physically impossible)
```{r}
sum(flow_data$flow <= 0, na.rm = TRUE)
```

### Extreme Values
```{r}
flow_data %>%
  arrange(desc(flow)) %>%
  select(date, flow, quality) %>%
  head(10)
```

### Duplicate Dates
```{r}
flow_data %>%
  count(date) %>%
  filter(n > 1)
```

### Summary Statistics
```{r}
flow_data %>%
  summarise(
    n_years     = n_distinct(year),
    n_obs       = n(),
    mean_flow   = round(mean(flow, na.rm = TRUE), 2),
    median_flow = round(median(flow, na.rm = TRUE), 2),
    sd_flow     = round(sd(flow, na.rm = TRUE), 2),
    max_flow    = round(max(flow, na.rm = TRUE), 2),
    p95_flow    = round(quantile(flow, 0.95, na.rm = TRUE), 2),
    p99_flow    = round(quantile(flow, 0.99, na.rm = TRUE), 2)
  )
```
---

## 3. Distribution
```{r}
ggplot(flow_data, aes(x = flow)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 80, fill = "blue", alpha = .5) +
  geom_density(colour = "black", linewidth = 0.8) +
  geom_vline(xintercept = mean(flow_data$flow), colour = "red") + 
  scale_x_log10() +
  theme_minimal() +
  labs(
    title    = "Distribution of Daily Streamflow",
    subtitle = "Log scale, Red Line = Mean",
    x        = "Flow (m³/s, log scale)",
    y        = "Density"
  )
```
---

## 4a. Seasonality by Month
```{r}
flow_data %>%
  group_by(month) %>%
  summarise(
    mean_flow   = mean(flow, na.rm = TRUE),
    median_flow = median(flow, na.rm = TRUE),
    p95_flow    = quantile(flow, 0.95, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols      = c(mean_flow, median_flow, p95_flow),
    names_to  = "statistic",
    values_to = "flow"
  ) %>%
  ggplot(aes(x = month, y = flow, colour = statistic)) +
  geom_line(linewidth = 0.8) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  theme_minimal() +
  labs(
    title  = "Seasonal Flow Patterns",
    x      = NULL,
    y      = "Flow (m³/s)",
    colour = NULL
  )
```

## 4b. Seasonal Flow Variation
```{r}
flow_data %>%
  mutate(season = factor(season, 
                         levels = c("Winter", "Spring", 
                                    "Summer", "Fall"))) %>%
  ggplot(aes(x = season, y = flow, fill = season)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3, outlier.size = .5) +
  scale_y_log10() +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    title    = "Flow Distribution by Season",
    subtitle = "Log scale",
    x        = NULL,
    y        = "Flow (m³/s, log scale)"
  )
```

---

## 5. Annual Maximum Trend
```{r}
flow_data %>%
  group_by(year) %>%
  summarise(max_flow = max(flow, na.rm = TRUE)) %>%
  ggplot(aes(x = year, y = max_flow)) +
  geom_point(colour = "blue", alpha = 0.8) +
  geom_smooth(method = "loess", colour = "red", se = TRUE, alpha = 0.1) +
  theme_minimal() +
  labs(
    title    = "Annual Maximum Flow Over Time",
    subtitle = "Red line = Loess Trend",
    x        = NULL,
    y        = "Annual Max Flow (m³/s)"
  )
```

---

## 6a. Autocorrelation
```{r}
acf_values <- acf(flow_data$flow, lag.max = 15, plot = FALSE)

data.frame(
  lag = as.numeric(acf_values$lag),
  acf = as.numeric(acf_values$acf)
) %>%
  ggplot(aes(x = lag, y = acf)) +
  geom_col(fill = "blue", alpha = 0.5, width = 0.7) +
  geom_hline(yintercept =  0.2, colour = "red",   linetype = "dashed") +
  geom_hline(yintercept = -0.2, colour = "red",   linetype = "dashed") +
  geom_hline(yintercept =  0,   colour = "black", linewidth = 0.4) +
  theme_minimal() +
  labs(
    title    = "Autocorrelation of Daily Flow",
    subtitle = "Red dashed lines = [-.2,.2] threshold",
    x        = "Lag (days)",
    y        = "ACF"
  )
```

High autocorrelation at short lags confirms that consecutive days of elevated flow are not independent, a single storm event produces several correlated days of high flow. This directly motivates the declustering step in the next notebook, where a minimum separation of 3 days between events is used to ensure independence before fitting the Poisson frequency model.

### 6b. Save Minimum Separation of 3 Days
```{r}
min_days_independent <- 3

saveRDS(min_days_independent, here::here("data", "clean", "min_days_independent.rds"))

```


---

## 7. Key Findings

**Distribution:**
Daily streamflow is strongly right skewed with mean > median. The spread between p95, mean, and median widens at higher flow levels meaning the gap between p95 and mean is substantially larger than the gap between mean and median, indicating increasing variance at higher percentiles.

**Seasonality:**
Flow peaks between February and mid-April driven by snowmelt. This pattern is consistent across mean, median, and p95. Right skewed behavior persists year-round implying its not driven by a single season.

**Annual maximum trend:**
The LOESS smoother reveals a non-monotonic trend. Maxima increased from ~30 m³/s in 1960 to ~50 m³/s by 1980, declined to ~40 m³/s through 2000, and have slowly increased to ~46 m³/s since. No strong linear trend but a modest post-2000 upward signal is visible.

**Autocorrelation:**
Strong autocorrelation at short lags as ACF drops below 0.2 at approximately 3 days. Consecutive high flow days are not independent, consistent with multi-day storm drainage behavior.

---

## 8. Modeling Implications

**GPD for severity:**
The right skewed heavy tailed distribution with increasing variance at higher percentiles directly supports GPD as the severity model. Extreme events are not just larger, they are disproportionately more variable. This is characteristic of heavy tailed behavior that GPD is designed to capture. The positive skew also suggests the shape parameter ξ is likely positive, consistent with the prior belief in Gamma(α, β) - 0.3.

**Poisson for frequency:**
The rare arrival of flood events combined with many daily opportunities satisfies the Poisson Limit Theorem conditions. However the strong short-lag autocorrelation confirms that raw daily exceedances are not independent, a declustering step is required before fitting the Poisson model. The ACF dropping below 0.2 at 3.5 days justifies the 3 day minimum separation criterion: conservative enough to ensure independence while retaining as many events as possible.

**Stationarity assumption:**
The non-monotonic LOESS trend and modest post-2000 upward signal suggest some non-stationarity in the historical record. However the absence of a strong monotonic trend means stationarity is a reasonable baseline assumption for Version 1. The observed trend provides empirical motivation for coupling the distributions' parameters to climate drivers in Version 2.

**Seasonal structure:**
The clear February to mid-April peak introduces seasonal variation in flood frequency that a stationary Poisson model ignores. This is a known limitation of Version 1, a seasonal frequency component where λ varies by season is a natural Version 2 extension.

