---
title: "Physical Threshold Selection - Don River"
author: "Dylan"
date: "`r Sys.Date()`"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


## Overview

This notebook documents how the physical flood threshold u_physical
was selected for the Don River at Todmorden (02HC024). u_physical is
the flow level in m³/s at which water leaves the Don River channel
and causes real world damage. It is used to define flood events for
the frequency model and as the lower bound for the Bayesian MRL
grid search in the next notebook.

The process was not straightforward. The initial approach attempted
to use the official TRCA flood warning threshold but was abandoned
due to an irreconcilable reference system mismatch between TRCA and
Environment Canada data. Three independent empirical approaches were
then used to converge on a defensible value. Each approach and its
limitations are documented transparently below.

---

## 1. Load Data
```{r}
library(tidyverse)
library(here)

flow_data    <- read_csv(here::here("data", "clean", "02HC024_clean.csv"))
rating_curve <- read_csv(here::here("data", "clean", "02HC024_rating_curve_clean.csv"))
```

---

## 2. Understanding The Water Level Data And The TRCA Problem

The Environment Canada water level data uses a local gauge datum,
not elevation above sea level. For 02HC024 Todmorden the normal
water level reads approximately **12.1m** in the EC data. TRCA's
real-time gauging system reports the same station using metres
above sea level (MASL):
```{r}
trca_levels <- tibble(
  level       = c("Normal", "Low Alarm", "High Alarm", "High-High Alarm"),
  water_level = c(77.59, 78.70, 79.00, 79.30),
  reference   = "TRCA Real-Time Gauging (MASL)"
)
trca_levels
```

The High Alarm level of **79.0m** MASL was the natural starting point
as TRCA defines this as the level at which active flooding occurs.
The implied datum offset of **65.4m** was estimated by taking the
difference between the two normal level readings: TRCA reports
77.59m MASL while Environment Canada records 12.21m for the same
normal conditions, giving 77.59 minus 12.21 equals 65.38m.
**Converting the High Alarm level to local datum then gives approximately 13.6m (79.0 minus 65.4), suggesting 14.0m** as a
reasonable working estimate for the flood threshold in EC units.
However this offset was estimated not verified against an official
benchmark elevation document, making 14.0m an approximation rather
than an authoritative value.

The water level data covers 2002 onward and has a maximum observed
value of `r round(max(rating_curve$water_level_m), 2)`m, which never
reached the estimated 14.0m flood threshold. This means the
threshold cannot be validated against observed discharge data.

All three approaches below therefore work directly from the
discharge record. A future improvement would be to contact TRCA
or Environment Canada directly to obtain the official benchmark
elevation for 02HC024, which would allow direct conversion and
independent validation of u_physical against the official warning
threshold.

---

## 3. Rating Curve — Water Level vs Discharge
```{r}
ggplot(rating_curve, aes(x = water_level_m, y = discharge_m3s)) +
  geom_point(alpha = 0.3, size = 0.8, colour = "steelblue") +
  geom_smooth(method = "loess", colour = "red",
              se = TRUE, alpha = 0.1) +
  geom_vline(xintercept = 14.0, colour = "black",
             linetype = "dashed") +
  annotate("text", x = 14.05, y = 60,
           label = "Estimated flood\nlevel (14.0m)",
           hjust = 0, size = 3) +
  annotate("rect", xmin = 13.55, xmax = 14.3,
           ymin = 0, ymax = Inf,
           alpha = 0.05, fill = "red") +
  theme_minimal() +
  labs(
    title    = "Rating Curve: Don River at Todmorden",
    subtitle = "Water level vs discharge (2002 onward). Shaded region = beyond observed data range.",
    x        = "Water Level (m, local datum)",
    y        = "Discharge (m³/s)"
  )
```

The rating curve shows a clear positive relationship between water
level and discharge. The estimated flood threshold of 14.0m lies
in the shaded region beyond the observed data range, making any
extrapolation unreliable.

---

## 4. Highest Observed Water Level Events
```{r}
rating_curve %>%
  arrange(desc(water_level_m)) %>%
  select(YEAR, month_name, DD, water_level_m, discharge_m3s) %>%
  head(20)
```

The highest observed water levels between 13.0 and 13.5m correspond
to discharge values of approximately 40 to 80 m³/s. The 2013 July
flood appears at 53.3 m³/s, consistent with what HYDAT records for
that date.

---

## 5. Approach 1: Extrapolation From Rating Curve
```{r}
rating_model <- lm(log(discharge_m3s) ~ water_level_m,
                   data = rating_curve)

predicted      <- predict(rating_model,
                          newdata = data.frame(water_level_m = 14.0))

(u_extrapolated <- round(exp(predicted), 1))
```

The log-linear extrapolation to 14.0m produces an estimated
discharge of `r u_extrapolated` m³/s, a physically impossible
result. The Amazon River flows at around 200,000 m³/s and
the Don River at its worst recorded event was approximately 400
to 500 m³/s. This confirms that extrapolating 0.5m beyond the
observed water level range is unreliable. Hence, approach 1 is a discarded.

---

## 6. Approach 2: Known Historical Flood Events

Toronto's Don River has a well documented flood history. The
following events were identified through Environment Canada
discharge records, IBC catastrophe loss reports, and the
Canadian Disaster Database:
```{r}
flood_events <- tibble(
  date   = as.Date(c(
    "1954-10-15",
    "1986-04-14",
    "2013-07-08",
    "2017-05-04",
    "2019-05-24"
  )),
  event  = c(
    "Hurricane Hazel",
    "Spring freshet flooding",
    "July thunderstorm — Toronto flash flood",
    "Spring freshet flooding",
    "Spring freshet flooding"
  ),
  type   = c(
    "Fluvial",
    "Fluvial",
    "Pluvial/Fluvial",
    "Fluvial",
    "Fluvial"
  ),
  source = c(
    "Canadian Disaster Database",
    "Environment Canada discharge record",
    "IBC — $940M insured loss (Ontario)",
    "Environment Canada discharge record",
    "Environment Canada discharge record"
  )
)

flood_events
```

Cross referencing these dates against the HYDAT flow data:
```{r}
confirmed_flood_flows <- flow_data %>%
  filter(date %in% flood_events$date) %>%
  filter(!is.na(flow)) %>%
  left_join(flood_events, by = "date") %>%
  select(date, event, type, flow)

confirmed_flood_flows
```

The 2013 event records a daily mean of
`r round(max(confirmed_flood_flows$flow), 1)` m³/s, the highest
value for any known flood date in the HYDAT record. The 2017 and
2019 values appear low. This is a consequence of using daily mean
discharge.(flood that peaked for 6 to 12 hours gets averaged
across the full 24 hour day, significantly understating the true
peak flow)


Hurricane Hazel (1954) is absent from the HYDAT comparison because
records for 02HC024 only begin in 1961. Hazel is historically
documented as the most catastrophic flood event on the Don River
with peak flows on the Humber River reaching approximately 1,700
m³/s. Its inclusion here acknowledges the existence of extreme tail
events beyond the observed record, directly relevant to the GPD
shape parameter discussion in the severity model notebook.

The minimum daily mean across all confirmed flood dates where HYDAT
data exists is `r round(min(confirmed_flood_flows$flow), 1)` m³/s
and the maximum is `r round(max(confirmed_flood_flows$flow), 1)`
m³/s. These are daily means, true instantaneous peaks were
substantially higher. u_physical should be below this minimum to
ensure all genuine flood events are captured while remaining above
normal flow conditions.

The 2013 event is classified as primarily **pluvial**: the discharge
recorded at the gauge reflects a combination of surface runoff and
riverine flow that cannot be separated at daily resolution. This
event is included for completeness and as an upper bound reference
but is not used as a primary anchor for u_physical given the
fluvial focus of this model. The threshold is derived primarily from the percentile analysis and sensitivity table in Sections 7 and 8.

---

## 7. Approach 3: Flow Percentiles
```{r}
p_flow_tbl <- flow_data %>%
  summarise(
    p90  = quantile(flow, 0.90,  na.rm = TRUE),
    p95  = quantile(flow, 0.95,  na.rm = TRUE),
    p99  = quantile(flow, 0.99,  na.rm = TRUE),
    p999 = quantile(flow, 0.999, na.rm = TRUE),
    max  = max(flow, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols      = everything(),
    names_to  = "percentile",
    values_to = "flow_m3s"
  ) %>%
  mutate(percentile = factor(percentile,
                             levels = c("p90", "p95", "p99",
                                        "p999", "max")))

ggplot(p_flow_tbl, aes(x = percentile, y = flow_m3s, group = 1)) +
  geom_line(colour = "blue", linewidth = 1) +
  geom_point(colour = "blue", size = 3) +
  geom_hline(yintercept = 30, colour = "red",
             linetype = "dashed") +
  annotate("text", x = 1.2, y = 32,
         label = "Initial estimate: 30 m³/s",
         colour = "red", size = 3, hjust = 0) +
  theme_minimal() +
  labs(
    title    = "Flow Percentiles — Don River at Todmorden",
    subtitle = "Exponential growth at upper tail confirms heavy tailed behavior",
    x        = "Percentile",
    y        = "Daily Mean Flow (m³/s)"
  )
```
```{r}
p_flow_tbl
```

The jump from p99 to p99.9 is nearly double the jump from p95 to
p99, and the jump from p99.9 to the maximum is larger still. This
accelerating growth at the upper tail is the visual signature of
heavy tailed behavior and directly supports GPD as the appropriate
severity model.

---

## 8. Threshold Decision

### Sensitivity Analysis

Event counts were checked across a range of candidate thresholds
before committing to 30 m³/s:
```{r}
sensitivity <- tibble(threshold = c(20, 25, 30, 35, 40)) %>%
  mutate(
    days_above   = map_int(threshold, ~sum(flow_data$flow >= .x,
                                           na.rm = TRUE)),
    pct_above    = round(days_above / nrow(flow_data) * 100, 2),
    rough_events = map_int(threshold, ~{
      flow_data %>%
        mutate(
          above     = flow >= .x,
          new_event = above & !lag(above, default = FALSE)
        ) %>%
        pull(new_event) %>%
        sum(na.rm = TRUE)
    })
  )

sensitivity
```

A threshold of 30 m³/s sits in the sweet spot - **original thought**... 

high enough to exclude normal elevated flow, low enough to retain sufficient
events for GPD fitting, and consistent with the onset of
flood-producing conditions observed in the rating curve and
historical event record.

### First Decision
```{r}
U_PHYSICAL <- 30
```

u_physical = `r U_PHYSICAL` m³/s is selected for the following
reasons. It sits above the p99 of `r round(quantile(flow_data$flow, 0.99, na.rm = TRUE), 1)` m³/s
meaning flows above 30 m³/s are rare, occurring on only
`r round(mean(flow_data$flow >= U_PHYSICAL, na.rm = TRUE) * 100, 2)`%
of days. It is consistent with observed flood events where all
documented events in the rating curve correspond to daily means of
40 to 80 m³/s, making 30 m³/s a conservative lower bound for
flood-producing flows. It provides sufficient events for modeling
with approximately 40 to 60 independent events expected after
proper declustering.

The known limitation is that u_physical is defined in terms of
daily mean discharge. True instantaneous flood peaks are higher.
This conservative threshold partially compensates by capturing
the onset of flood conditions rather than peak conditions.

Note that the event count here is a rough upper bound based on
threshold crossings, not proper declustering. The full
declustering algorithm using a 3 day minimum separation is
implemented in notebook 03_event_extraction.Rmd.


### IMPORTANT CHANGE On u_physical After Declustering Analysis

After implementing the declustering function the following event
characteristics were observed at u_physical = 30 m³/s:

```{r}

invisible(source(here::here("R_functions", "declustering.R"), echo = FALSE))

flood_events_30 <- decluster_events(
  flow_data            = flow_data,
  threshold            = 30,
  min_days_independent = 3
)

flood_events_30 %>%
  summarise(
    n_events        = n(),
    mean_duration   = mean(duration),
    median_duration = median(duration),
    max_duration    = max(duration)
  )

flood_events_30 %>%
  arrange(start_date) %>%
  mutate(days_since_last = as.numeric(start_date - lag(end_date))) %>%
  summarise(
    mean_gap    = mean(days_since_last, na.rm = TRUE),
    median_gap  = median(days_since_last, na.rm = TRUE),
    min_gap     = min(days_since_last, na.rm = TRUE),
    pct_under_3 = mean(days_since_last < 3, na.rm = TRUE) * 100
  )
```

The median event duration of 1 day confirms that the Don River
responds and recovers rapidly at 30 m³/s, consistent with a
small urban catchment with fast drainage. The minimum inter-event
gap of 4 days and zero events separated by fewer than 3 days
confirms all 129 events are independent. The 3 day
separation criterion was therefore conservative for this dataset as natural event separation was already occurring.

However the 1 day median duration brings a physical concern.
A single day spike barely above 30 m³/s is unlikely to represent
a damaging flood event. The rating curve analysis in
Section 4 showed all documented flood events correspond to daily
means of 40 to 80 m³/s. Events between 30 and 40 m³/s likely
represent elevated flow conditions rather than active flooding.

Comparing event characteristics across candidate thresholds:
```{r}
threshold_comparison <- tibble(threshold = c(30, 35, 40, 45, 50)) %>%
  mutate(
    n_events = map_int(threshold, \(u) {
      decluster_events(flow_data, u, 3) %>% nrow()
    }),
    median_duration = map_int(threshold, \(u) {
      events <- decluster_events(flow_data, u, 3)
      as.integer(median(events$duration))
    }),
    mean_Q_peak = map_dbl(threshold, \(u) {
      events <- decluster_events(flow_data, u, 3)
      round(mean(events$Q_peak), 1)
    })
  )

threshold_comparison
```

Raising u_physical to 40 m³/s produces 54 independent events
with a higher mean Q_peak of 53.4 m³/s, better aligned with
the documented flood damage range of 40 to 80 m³/s observed
in the rating curve. The median event duration remains 1 day
across all candidate thresholds, a consistent characteristic
of the Don River as a fast responding urban catchment rather
than a product of the threshold choice.

54 events sits at the boundary of the practical minimum of 50
exceedances recommended by Coles (2001) for reliable GPD parameter estimation. The Bayesian framework partially relaxes this requirement through prior
regularization, but 40 m³/s is the highest threshold that
remains statistically workable. Raising to 45 m³/s drops event
count to 35, below the minimum regardless of framework.

u_physical is revised from 30 to 40 m³/s on **physical grounds**.
At **30 m³/s** the mean Q_peak of 42.2 m³/s and **single day median duration suggest many events represent brief elevated flow rather than genuine flooding**. At **40 m³/s** the mean Q_peak of 53.4 m³/s is consistent with the onset of documented flood
damage and the threshold sits **above the p99.9 of 52.5 m³/s**,
capturing only genuinely extreme events.

## 9. Save u_physical
```{r}
U_PHYSICAL <- 40

flood_events <- decluster_events(
  flow_data            = flow_data,
  threshold            = U_PHYSICAL,
  min_days_independent = 3
)

saveRDS(U_PHYSICAL,
        here::here("data", "clean", "u_physical.rds"))

write_csv(flood_events,
          here::here("data", "clean", "flood_events.csv"))
```

u_physical updated to `r U_PHYSICAL` m³/s and saved. The flood
events table has been updated accordingly with
`r nrow(flood_events)` independent events.

---

## Summary

| u_physical | 40 m³/s | Above p99.9, mean Q_peak consistent with documented flood damage, 54 independent events above GPD minimum |