---
title: "Posterior Predictive Checks — Don River Flood Risk Model"
author: "Dylan Thackeray"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Setup

```{r}
library(tidyverse)
library(here)
library(bayesplot)
library(posterior)
library(loo)
library(evd)

set.seed(1234)

# Observed data
severity_df <- read_csv(here::here("data", "model", "severity_data.csv"))
freq_df     <- read_csv(here::here("data", "model", "frequency_data.csv"))

exceedances <- severity_df$exceedance
event_counts <- freq_df$n_events

# Fitted models
fit_pois <- readRDS(here::here("data", "model", "bayesian_model_fits", "poisson_fit.rds"))
fit_gpd  <- readRDS(here::here("data", "model", "bayesian_model_fits", "gpd_fit.rds"))

# Posterior draws
pois_draws <- as_draws_df(fit_pois$draws())
gpd_draws  <- as_draws_df(fit_gpd$draws())
```

# 1. Trace Plots (MCMC Convergence)

```{r fig.height=3}
mcmc_trace(fit_pois$draws(), pars = "lambda") +
  labs(title = "Trace plot — lambda (Poisson frequency)")

mcmc_trace(fit_gpd$draws(), pars = c("sigma", "xi")) +
  labs(title = "Trace plots — sigma and xi (GPD severity)")
```

All chains mix well. R-hat values:

```{r}
fit_pois$summary("lambda")[, c("variable", "mean", "sd", "q5", "q95", "rhat", "ess_bulk")]
fit_gpd$summary(c("sigma", "xi"))[, c("variable", "mean", "sd", "q5", "q95", "rhat", "ess_bulk")]
```

# 2. Prior vs Likelihood vs Posterior

This plot is constructed analytically (not from `generated_quantities`) to show how the data updates each prior.

```{r fig.height=4}
lambda_vals <- seq(0.3, 2.5, length.out = 300)
sigma_vals  <- seq(5, 30, length.out = 300)
xi_vals     <- seq(-0.6, 0.5, length.out = 300)

sigma_fix <- mean(gpd_draws$sigma)
xi_fix    <- mean(gpd_draws$xi)

# Likelihoods (profile, fixing other parameters at posterior mean)
log_lik_lambda <- sapply(lambda_vals, \(l) sum(dpois(event_counts, l, log = TRUE)))
log_lik_sigma  <- sapply(sigma_vals,  \(s) sum(dgpd(exceedances, loc = 0, scale = s,   shape = xi_fix, log = TRUE)))
log_lik_xi     <- sapply(xi_vals,     \(x) sum(dgpd(exceedances, loc = 0, scale = sigma_fix, shape = x, log = TRUE)))

# Normalize to density scale
lik_lambda <- exp(log_lik_lambda - max(log_lik_lambda))
lik_sigma  <- exp(log_lik_sigma  - max(log_lik_sigma))
lik_xi     <- exp(log_lik_xi     - max(log_lik_xi))

# Priors
prior_lambda <- dgamma(lambda_vals, shape = 2, rate = 2)
prior_sigma  <- dgamma(sigma_vals,  shape = 2, rate = 0.1)
prior_xi     <- dnorm(xi_vals, mean = -0.1, sd = 0.2)

# Posterior kernel densities
post_lambda <- density(pois_draws$lambda)
post_sigma  <- density(gpd_draws$sigma)
post_xi     <- density(gpd_draws$xi)

# Build data frames
df_all <- bind_rows(
  tibble(x = lambda_vals, Prior = prior_lambda / max(prior_lambda),
         Likelihood = lik_lambda, Parameter = "lambda"),
  tibble(x = sigma_vals,  Prior = prior_sigma  / max(prior_sigma),
         Likelihood = lik_sigma,  Parameter = "sigma"),
  tibble(x = xi_vals,     Prior = prior_xi     / max(prior_xi),
         Likelihood = lik_xi,     Parameter = "xi")
) %>%
  pivot_longer(c(Prior, Likelihood), names_to = "Type", values_to = "Density")

post_df <- bind_rows(
  tibble(x = post_lambda$x, Density = post_lambda$y / max(post_lambda$y), Parameter = "lambda"),
  tibble(x = post_sigma$x,  Density = post_sigma$y  / max(post_sigma$y),  Parameter = "sigma"),
  tibble(x = post_xi$x,     Density = post_xi$y     / max(post_xi$y),     Parameter = "xi")
)

ggplot() +
  geom_line(data = df_all, aes(x, Density, color = Type), linewidth = 1.1) +
  geom_line(data = post_df, aes(x, Density), color = "black", linewidth = 1.1) +
  facet_wrap(~Parameter, scales = "free") +
  scale_color_manual(values = c("Prior" = "steelblue", "Likelihood" = "tomato")) +
  labs(
    title    = "Prior, Likelihood, and Posterior",
    subtitle = "Black = posterior (Stan). Densities normalized to [0,1] for visual comparison.",
    x = NULL, y = "Normalized density"
  ) +
  theme_minimal()
```

For λ and σ, the likelihood is concentrated and the posterior closely follows it — both parameters are well-identified. For ξ, the likelihood is relatively flat, so the posterior remains wide and reflects meaningful prior influence. This is expected with 54 exceedances.

# 3. Posterior Predictive Check — Frequency

`n_rep` is drawn directly from Stan's `generated_quantities` block and reflects genuine posterior predictive uncertainty.

```{r}
# Extract n_rep matrix: rows = draws, cols = years
n_rep <- fit_pois$draws("n_rep", format = "matrix")

# bayesplot PPC: observed counts vs 50 posterior predictive replications
ppc_bars(
  y    = as.integer(event_counts),
  yrep = n_rep[sample(nrow(n_rep), 50), ]
) +
  labs(
    title    = "PPC — Annual Event Counts",
    subtitle = "Dark = observed. Light = posterior predictive replications."
  )
```

```{r}
# Mean events/year: observed vs posterior predictive distribution
ppc_stat(
  y    = as.integer(event_counts),
  yrep = n_rep,
  stat = "mean"
) +
  labs(title = "PPC — Mean Events per Year", x = "Mean count")
```

# 4. Posterior Predictive Check — Severity

`y_rep` is drawn directly from Stan's `generated_quantities` block.

```{r}
# Extract y_rep matrix: rows = draws, cols = observations
y_rep <- fit_gpd$draws("y_rep", format = "matrix")

# Density overlay: observed vs 50 posterior predictive replications
ppc_dens_overlay(
  y    = exceedances,
  yrep = y_rep[sample(nrow(y_rep), 50), ]
) +
  labs(
    title    = "PPC — Exceedance Density",
    subtitle = "Dark = observed. Light = 50 posterior predictive replications."
  )
```

```{r}
# ECDF overlay on log scale: sensitive to tail behavior
ppc_ecdf_overlay(
  y    = exceedances,
  yrep = y_rep[sample(nrow(y_rep), 50), ]
) +
  scale_x_log10() +
  labs(
    title    = "PPC — Exceedance ECDF (log scale)",
    subtitle = "Divergence in upper tail reflects ξ uncertainty, not model failure."
  )
```

```{r}
# Distribution of replicated maximums vs observed maximum
ppc_stat(
  y    = exceedances,
  yrep = y_rep,
  stat = "max"
) +
  geom_vline(xintercept = max(exceedances), linetype = "dashed") +
  labs(
    title    = "PPC — Distribution of Simulated Maximums",
    subtitle = "Dashed = observed maximum (68 m³/s above threshold).",
    x        = "Maximum exceedance (m³/s)"
  )
```

```{r}
# 95th percentile
ppc_stat(
  y    = exceedances,
  yrep = y_rep,
  stat = function(x) quantile(x, 0.95)
) +
  labs(title = "PPC — 95th Percentile of Exceedances", x = "p95 exceedance (m³/s)")
```

# 5. LOO Cross-Validation

```{r}
log_lik_pois <- fit_pois$draws("log_lik", format = "matrix")
log_lik_gpd  <- fit_gpd$draws("log_lik",  format = "matrix")

loo_pois <- loo(log_lik_pois)
loo_gpd  <- loo(log_lik_gpd)

print(loo_pois)
print(loo_gpd)
```

```{r}
# Pareto-k plot: identifies observations with outsized influence on GPD posterior
plot(loo_gpd, diagnostic = "k", label_points = TRUE) +
  labs(
    title    = "LOO Pareto-k Diagnostics — GPD Severity Model",
    subtitle = "k > 0.7 indicates an influential observation. k > 0.5 warrants attention."
  )
```

Observations with k > 0.5 correspond to the largest exceedances. These are the data points that most heavily influence the GPD shape parameter (ξ) posterior and are expected in extreme value models.

# 6. Summary

The posterior predictive checks confirm the following:

- **Frequency model:** The Poisson model reproduces the observed annual count distribution well. Mean events per year is recovered within posterior predictive uncertainty.

- **Severity model — central behavior:** The density overlay shows good agreement in the bulk of the exceedance distribution. σ is well-identified.

- **Severity model — tail behavior:** The ECDF and maximum-statistic checks show that the observed maximum falls comfortably within the posterior predictive distribution, but the simulated tail is substantially wider than the observed data. This reflects posterior uncertainty in ξ, not a model failure. With 54 observations, ξ is weakly identified and the model honestly propagates this into tail projections.

- **LOO:** Pareto-k values confirm that the largest exceedances are the most influential observations — consistent with EVT. No observations have k > 0.7 (which would indicate a numerical reliability problem with the LOO estimate).
